\plaintitle

### 2025-06-04 docs/style/article-body.tex
the draft sent in for preliminary supervision

\section{ABSTRACT}\label{abstract}

True random number generation is indispensable for a wide range of cryptographic and distributed computing applications, yet traditional pseudo-random number generators remain deterministic, and thus vulnerable to state‐recovery attacks. Quantum random number generation (\emph{QRNG}) offers a robust alternative by harnessing inherent quantum fluctuations -- in particular, optical QRNG (\emph{OQRNG}) based on shot noise provides an accessible source of entropy from which true random numbers can be generated. However, existing OQRNG solutions typically rely on expensive or bespoke hardware and defer post‐processing to host computers, limiting portability and integration. In this work, we investigate the feasibility of performing Toeplitz randomness extraction directly on resource constrained microcontrollers, thus creating a fully self‐contained QRNG system. We address two primary research areas: RA1 developing an implementation of Toeplitz extraction that minimizes processing time on embedded hardware, and RA2 ensuring that the extraction algorithm does not become the throughput bottleneck compared to ADC conversion or USB transfer speeds.

Our results demonstrate that, by eliminating dynamic allocations and leveraging bit parallel techniques, Toeplitz extraction on 64-bit inputs can be executed in approximately \(0.05 \mu s\) on the Teensy 4.1 -- well below the \(2.667 \mu s\) threshold imposed by a 12-bit, 2 MSPS ADC. Even accounting for variance in individual measurements, these speeds confirm that the algorithmic implementation does not constrain overall throughput. On the Pico 2, the same implementations yield sub‐microsecond performance, albeit with a wider variance. However, our most efficient implementations are limited to 64‐bit inputs -- extending to larger input bit sizes incurs significant overhead due to the absence of wider integer types in typical MCU toolchains.

In conclusion, Toeplitz extraction can be performed in real time on low‐power microcontrollers without becoming the system's performance bottleneck, validating the viability of a portable QRNG thumbstick. Future work should explore custom data types or lightweight libraries for 128- and 256-bit size extraction, as well as integration with a live OQRNG device to assess end‐to‐end randomness quality and throughput.

@@

There will unequivocally be a bottleneck for the processing speed. For instance, the speed at which the ADC can process the optical signal into raw bits as well as the speed that the USB output can transfer processed random number to the host computer will be limiting factors. Further details on the limitations of the ADC will be outlined in Section 3. The slowest of these bottlenecks will inevitably be the limiting factor for any implementation. Our research aims to ensure that our implementation of Toeplitz extraction does not become the limiting factor, but rather processing data fast enough to match or exceed the speed of the hardware.

Section 2 of this article will introduce the theory that allows for QRNG, and how this will be utilized in our works. Section 3 delves further into the hardware and algorithms our work will use, with related works in optimizing Toeplitz extraction listed under section 4. Section 5 will present our methodology and implementation strategy, as well as some limitations imposed on our work. Finally, section 7 will present the results of our experimentation.

@@

Our work is a practical continuation of the work of Clason \cite{Clason2023}. In this work, quantum shot noise originating from photodiodes was studied, and in so doing a prototype device was constructed. This prototype utilized an LED soldered millimeters apart from a photodiode, which in turn outputs a variable analog signal from the device. In this section, we introduce the remaining hardware used for our implementations as well as the considerations taken in order to shift the focus from physics to computer science.

@@

As described in Clasons work \cite{Clason2023}, the OQRNG-device is an electro-optical system which measures optical shot noise, generating quantum randomness. The device has an LED and a photodiode positioned a few millimeters apart, ensuring efficient light coupling. The photodiode detects light from the LED, and converts the light into a current signal, which is sent to a transimpedance amplifier to convert it into measurable voltage. In order to minimize disruptions by other external lights, the system is enclosed in a shielded measurment box.

@@

Microcontrollers (\emph{MCUs}) are compact and low-power computing devices designed for embedded systems and real-time operations, and suitable as a processing unit for the purposes of this work. Unlike general CPUs, an MCU integrates a processor, memory and peripherals (\emph{such as an ADC}) into a single chip. Furthermore, modern MCUs often feature advanced microarchitectural elements to enhance processing capabilities on single threads (\emph{such as dual-issue superscalar architectures, allowing the MCU to run several instructions per CPU cycle}), making them suitable candidates for the post-processing required for OQRNG-data.

Since MCUs often function under strict timing requirements, it is critical to have effective ways to access memory and transfer data for processing in real time. High-performance MCUs enhance memory usage in various ways. Some of them use Tightly Coupled Memory\footnote{\url{https://www.farnell.com/datasheets/1913106.pdf}} (\emph{TCM}), which gives fast SRAM with specific access routes for important data, avoiding cache misses and guaranteeing consistent performance. Moreover, instruction and data caching techniques, including instruction pre-fetching and branch prediction, help minimize execution delays in computationally intensive real-time applications. Another important aspect is Direct Memory Access\footnote{\url{https://www.sciencedirect.com/topics/computer-science/direct-memory-access}} (DMA), which enables data transfer between peripherals such as the ADC and RAM, without CPU intervention. This offloading reduces processing overhead, allowing the MCU to manage fast data transfers effectively. These improvements are especially significant for Toeplitz extraction, where large amount of random data needs to be processed and sent quickly with low delays. Efficient memory management guarantees that randomness extraction can occur rapidly without major slowdowns in computing. Both approaches will be tested during development.

@@

In our work, we intend to use Teensy 4.1\footnote{\url{https://www.pjrc.com/store/teensy41.html}} based on the ARM Cortex-M7 pictured to the right in fig.~\ref{fig:mcu}. This MCU is especially suitable for computationally demanding tasks involving randomness extraction due to its dual-issue superscalar architecture and Digital Signal Processing (\emph{DSP}) capabilities. The floating-point unit\footnote{\url{https://www.sciencedirect.com/topics/computer-science/floating-point-unit}} (\emph{FPU}) and Single Instruction, Multiple Data (\emph{SIMD}) style DSP instructions improve how quickly it can perform bitwise and arithmetic tasks, which is crucial for quick Toeplitz extraction. SIMD-controlled DSP architectures, as described by Han et al. \cite{simd-dsp}, leverage parallel vectorized computation to accelerate matrix operations -- making them highly effective for Toeplitz matrix-vector multiplications.

In order to evaluate how efficient our implementation can become, our aim is to try our implementation on other MCU with different level of power and hardware support. Whereas Teensy 4.1, operating at \texttt{600\ MHz}, is our primary development platform which we will evaluate closely, we aim to run our implementations on the Raspberry Pi Pico 2\footnote{\url{https://datasheets.raspberrypi.com/pico/pico-2-product-brief.pdf}}, operating at \texttt{150\ MHz} (\_pictured to the right in fig.~\ref{fig:mcu_}). Due to the lower computational power of this MCU, there may be significant issues in utilizing this weaker model, yet it is significantly cheaper and easier to access. Testing will consist solely of running the implementation on these controllers and measuring execution speed and correctness of the output.

@@

The raw bits from the ADC can potentially have some deterministic patterns, and as such have to be processed somehow in order to remove these patterns. Several methods exist for this purpose, and for our work, we will perform this pre-processing via Toeplitz extraction. The main focus of this study is to implement this extraction algorithm as effectively as possible on resource constrained hardware.

@@

The main focus of this work is implementing this algorithm as efficiently as possible on our MCUs, and as such, several optimization efforts need to be taken into account during our experimentation.

@@

With the assumption that the OQRNG-device produces a truly random analog signal, we can now clearly define the scope in which this thesis operates. Considering the maximum conversion speed from the ADC and the USB-output from the MCU, we have a clear bound over 24 Mbit/s (\emph{imposed by the ADC}) in which Toeplitz extraction needs to be executed. Any speeds over 2.86 MB/s allows us to upgrade the ADC iteratively to continue increasing the output speed. Any implementation of Toeplitz extraction must then execute fast enough on any given microcontroller feasible for the proposed quantum RNG-thumbstick as to not be the decisive limiting factor.

@@

Another important point that is often neglected in randomness extraction using Toeplitz matrices is how the seed key is handled when forming the Toeplitz matrix. Numerous systems, such as those developed by Chouhan et al. \cite{toeplitz-desc} and Zhang et al. \cite{toeplitz}, depends on fixed seeds. Fixed seeds can create security risks over prolonged operation time. To tackle this problem, Lin et al \cite{lin} proposed a method for seed-renewable Toeplitz post-processing in QRNG. Their strategy incorporates a dynamic seed pool within the FPGA, where each instance of post-processing picks a new, randomly selected seed. Thus, minimizing temporal correlations between extractions. Furthermore, an external seed updating mechanism via Peripheral Component Interconnect Express (\emph{PCIe}) ensures that seeds are refreshed whenever a certain security limit is reached. Compared to fixed-seed methods, this renewable approach enhances cryptographic robustness and ensures sustained high-security randomness extraction in real-world applications.

Efficient Toeplitz matrix-vector multiplication (\emph{TMVM}) is critical for optimizing randomness extraction that relies on Toeplitz, particulary in constrained hardware environments. Liao et al. \cite{liao} showed that this process could be greatly accelerated using Fast Fourier Transform (\emph{FFT}) and its inverse (\emph{IFFT}) -- reducing computational complexity from \(O(n^2)\) to \(O(n \log n)\). Their implementation on FPGA utilized this approach for deep neural networks, resulting in a 28.7 times decrease in model size while still achieving fast inference speeds. By using FFT and IFFT acceleration, Toeplitz post-processing for randomness extraction could achieve higher throughput. Thus, potentially could improve performance and reduce latency.

@@

\textbf{Iteration 3 - Batching and Hardware optimization:} This iteration focused on optimizing performance through batching and the use of ARM-native instructions, beginning by testing batching alone. This was followed by isolated use of ARM instructions such as \texttt{\_\_builtin\_popcountll()} (\emph{which counts the number of set bits in an unsigned integer}). After establishing their individual effect, we combined both techniques, Multiple batch sizes were tested to determine their impact. More details and benchmarks for each configuration can be found in the section 6.

@@

Phase two focused on addressing inefficiencies and design issues that were unintentionally introduced or not addressed during earlier iterations. Rather than continuing with new algorithmic ideas, this phase aimed to identify and fix structural problems. Several assumptions from phase one were re-evaluated -- such as the benefits of certain data structures or abstractions.

@@

The way \texttt{ent} calculates the entropy of a given binary file is by checking the frequency of certain byte values. If some byte values are more frequent, the binary file has a low entropy -- essentially singifying that there are predictable patterns. Conversely, if all byte values appear roughly equally, the binary file has a high entropy -- meaning that it is more random, and less predictable. The maximum entropy of any given file is \(\log_2(256) = 8 \text{ bits per byte}\), meaning that a perfectly random binary file has an entropy score of 8. For the purposes of our work, we consider values greater than \(7.99\) to be acceptable.

Furthermore, \texttt{ent} also provides a measurement of the possible compression ratio of the file, e.g.~how much more the file can be theoretically compressed. This is calculated as \(1 - \frac{\text{entropy}}{\text{maximum entropy}}\), signifying that the file cannot be compressed further -- that the data is already as compressed as its random distribution allows. Essentially, this informs us that the file has no more detectable statistical patterns or compressable regularities remaining. We consider any file where \texttt{ent} reports a compression ratio of \(0\%\) to be acceptable.

In regards to execution time, Hyncica et. al. \cite{micromeasurements} propose that measuring execution time of algorithms directly via the microcontrollers internal timers (while subtracting the interrupt overhead) provides adequate measurements of the execution speed of an algorithm. An additional advantage is that the same code can be used to measure execution speed on several different microcontrollers, rather than relying on counting CPU cycles (as the process for this may vary greatly between controllers).

@@

To isolate the execution speed of the algorithm as precisely as possible, we only intend to measure the execution speed of the algorithm. Whereas there is some other code present in each iteration (\emph{primarily packing the output data to be sent to the host computer in a workable format}), the brunt of the work should reasonably be performed by the algorithm. As such, we expect the time spent on executing other code to be negligible. Furthermore, this code performs significantly less work per cycle, and does not exponentially increase the larger the input and output requirements increase.

@@

In section 6, this calculation will be used to derive the execution speed of the various iterations.

@@

We created a script to facilitate easier testing, which is attached to this paper. Using this script, all bitstrings used for testing can be evaluated against the baseline, ensuring that the output from the new iteration matches the baseline exactly. Furthermore, the script also provides the average execution time of only the Toeplitz extraction in microseconds. A brief overview of the architecture of this test script can be seen in Figure \ref{fig:testscript-architecture}.

@@

Table \ref{tab:iter0} presents the average execution time of the naive implementation (\emph{iteration 0}) across different input sizes for both the Teensy 4.1 and the Raspberry Pi Pico 2. The results reveal that the Teensy consistently achieves lower execution times compared to the Pico, with performance differences becoming more pronounced as thje bit size increases. This behaivor aligns with the expected difference in hardware specifications, such as CPU clock frequency and memory bandwidth.

@@

Additionally, the isolated effect of applying a single bitmask operation \texttt{\&\ 1} was evaluated. This resulted in a slight reduction in average execution time, from \texttt{13.1564} \(\mu s\) to \texttt{13.1000} \(\mu s\), although this specific result is not included in the tables above.

@@

Table \ref{tab:iter3} presents the average execution time of iteration 3 on the Teensy 4.1 for varying batching and bit sizes. Additionally, when performing the same 1024/1024-bit operation, the Raspberry Pi Pico 2 showed average execution time of \texttt{80806.006} \(\mu s\) indicating a substantially lower throughput compared to the Teensy.

@@

Table \ref{tab:iter4} presents the execution time of iteration 4, which utilizes a 4x unrolled bit-processing loop. Compared to erlier iterations, this approach yields a substantial performance improvments on both the Teensy and the Raspberry Pi Pico Mcu's.

@@

An additional fully unrolled variant, where loops were entirely eliminated, produced a measured execution time of \texttt{0.0491} \(\mu s\). However, this result was later deemed invalid due to packaging error in the implementation. Specifically, the algorithm was incorrectly designed to transmit only a single output bit instead of full 32-bit expected output. This lead to an underestimation of the true execution time, as the cost of correctly formatting and sending the full result was omitted from measurement.

@@

Table \ref{tab:iter5} presents the results of iteration 5, in which vector structure were removed in favor of fixed-size integer types \texttt{uint32\_t} and \texttt{uint64\_t}. This iteration led to an extremely efficient implementation that approaches the physical execution limits of the Teensy 4.1. Due to the removal of \texttt{std::vector}, this implementation is limited by the available integer sizes of the MCU, and could as such only be tested with at most 64 bits. The ramifications of this will be discussed in more detail in Section 7. However, the results remain inconsistent---while the average is low, individual measurements vary, indicating that the implementation does not yet achieve stable execution at the architectural limit.

@@

Table \ref{tab:iter6} presents the results of iteration 6 using 64 bit size, which was introduced to address the structural limitations encountered in iteration 5. Specifically, the fixed-width output constrains imposed by the removal of vector. Several alternative data structures were evaluated, with the original vecotr-based implementation from iteration 1 serving as a baseline.

Notably, while the bitset-based implementation achieved an average execution time of 0.0474\,\(\mu s\), this value represents an average and not a consistently stable result. Individual measurements showed significant variance. This suggests that, despite nearing the theoretical limits outlined in Section 3.3, the implementation has not yet reached a consistently minimal execution time. Additional tests at 128-bit input size yielded 179.8712\,\(\mu s\), confirming that the system still operates within the physical capabilities of the Teensy 4.1.

@@

Observing the results, we can place them into the context of the limits imposed by the ADC, as discussed in Section 3.2. In equation (5) (\emph{as seen in Section 5.3}), we calculate the average execution speed required to be \(2.667 \mu s\) for the ADC that requires soldering to the microcontroller. From Tables \ref{tab:iter5} and \ref{tab:iter6} we demonstrate execution speeds well below that, with iteration 5 as well as iteration 6 -- specifically with bitsets -- being the most prominent results to discuss. Here, our implementation achieved speeds of \(0.0501 \mu s\) and \(0.0474 \mu s\) respectively which within the bounds of the ADC listed in Section 3.2. The built in ADC on Teensy has an effective sample rate of 1 MSPS (\emph{e.g.~half of MAX11102AUB}), which would require an average execution speed of \textasciitilde{}\(2.667/2 \mu s \approx 1.334 \mu s\), making our implementation suitable even for that limited conversion speed.

@@

However, it is important to note that both of these implementations has a hard limit of working on 64 bits input. Whereas larger bitstrings still yield results with high entropy, the execution speed increases far above the hardware limitations, as evident in \ref{tab:iter6}. Iteration 5 is hard capped at using only 64 bits input, as we need to store the bits in fixed-size integers before running them through the algorithm, and on embedded machines we have no larger integer types available.

One issue still remains concerning the size of the input string and subsequent output string. The larger the size of the matrix provided to the Toeplitz function (\emph{e.g.~the larger our input}), the higher the potential overall entropy becomes. As our implementation is essentially capped at 64 bits input at this point in time (\emph{without being limited by the hardware}), this may or may not prove to not be secure enough for use in real world applications. This, however, is out of scope for this particular thesis, but a consideration nonetheless.

@@

As mentioned in Section 7, this comes at a price. The implementation that did indeed perform above expectation could only output a 32-bit integer, as we cannot take any larger input than 64 bits without any more complex data structure. How much of a flaw this is for a system utilizing this method of OQRNG is beyond the scope of this thesis to investigate. However, it is likely that a larger output is required for any application revolving around tight security. For less critical systems and less essential functionality, this method might, however, prove just secure enough. For instance, an IoT-device that requires a random number output for symmetric cryptography when connecting to a user device may be an ideal candidate for this method of OQRNG.

>>>>

@@

There will unequivocally be a bottleneck for the processing speed. For instance, the speed at which the ADC can process the optical signal into raw bits as well as the speed that the USB output can transfer processed random number to the host computer will be limiting factors. Further details on the limitations of the ADC will be outlined in Section 3.2. The slowest of these bottlenecks will inevitably be the limiting factor for any implementation. Our research aims to ensure that our implementation of Toeplitz extraction does not become the limiting factor, but rather processing data fast enough to match or exceed the speed of the hardware.

Section 2 of this article will introduce the theory that allows for QRNG, and how this will be utilized in our works. Section 3 delves further into the hardware and algorithms our work will use, with related works in optimizing Toeplitz extraction listed under section 4. Section 5 will present our methodology and implementation strategy, as well as some limitations imposed on our work. Finally, Section 7 will present the results of our experimentation, Section 8 presents our conclusions and Section 9 presents future work.

@@

Our work is a practical continuation of the work of Clason \cite{Clason2023}. In this work, quantum shot noise originating from photodiodes was studied, and in so doing a prototype device was constructed. In this section, we introduce the remaining hardware used for our implementations as well as the considerations taken in order to shift the focus from physics to computer science.

@@

As described in Clasons work \cite{Clason2023}, the OQRNG-device is an electro-optical system which measures optical shot noise, generating quantum randomness. The device has an LED and a photodiode positioned a few millimeters apart, ensuring efficient light coupling. The photodiode detects light from the LED, and converts the light into a current signal, which is sent to a transimpedance amplifier to convert it into measurable voltage. In order to minimize disruptions by other external lights, the system is enclosed in a shielded measurement box.

@@

Microcontrollers (\emph{MCUs}) are compact and low-power computing devices designed for embedded systems and real-time operations, and suitable as a processing unit for the purposes of this work. Unlike general CPUs, an MCU integrates a processor, memory and certain peripherals (\emph{such as an ADC}) into a single chip. Furthermore, modern MCUs often feature advanced microarchitectural elements to enhance processing capabilities on single threads (\emph{such as dual-issue superscalar architectures, allowing the MCU to run several instructions per CPU cycle}), making them suitable candidates for the post-processing required for OQRNG-data.

Since MCUs often function under strict timing requirements, it is critical to have effective ways to access memory and transfer data for processing in real time. High-performance MCUs enhance memory usage in various ways. Some of them use Tightly Coupled Memory\footnote{\url{https://www.farnell.com/datasheets/1913106.pdf}} (\emph{TCM}), which gives fast SRAM with specific access routes for important data, avoiding cache misses and guaranteeing consistent performance. Moreover, instruction and data caching techniques, including instruction pre-fetching and branch prediction, help minimize execution delays in computationally intensive real-time applications. Another important aspect is Direct Memory Access\footnote{\url{https://www.sciencedirect.com/topics/computer-science/direct-memory-access}} (\emph{DMA}), which enables data transfer between peripherals such as the ADC and RAM, without CPU intervention. This offloading reduces processing overhead, allowing the MCU to manage fast data transfers effectively. These improvements are especially significant for Toeplitz extraction, where large amount of random data needs to be processed and sent quickly with low delays. Efficient memory management guarantees that randomness extraction can occur rapidly without major slowdowns in computing. Both approaches will be tested during development.

@@

In our work, we intend to use Teensy 4.1\footnote{\url{https://www.pjrc.com/store/teensy41.html}} based on the ARM Cortex-M7 pictured to the left in fig.~\ref{fig:mcu}. This MCU is especially suitable for computationally demanding tasks involving randomness extraction due to its dual-issue superscalar architecture and Digital Signal Processing (\emph{DSP}) capabilities. The floating-point unit\footnote{\url{https://www.sciencedirect.com/topics/computer-science/floating-point-unit}} (\emph{FPU}) and Single Instruction, Multiple Data (\emph{SIMD}) style DSP instructions improve how quickly it can perform bitwise and arithmetic tasks, which is crucial for quick Toeplitz extraction. SIMD-controlled DSP architectures, as described by Han et al. \cite{simd-dsp}, leverage parallel vectorized computation to accelerate matrix operations -- making them highly effective for Toeplitz matrix-vector multiplications.

In order to evaluate how efficient our implementation can become, our aim is to try our implementation on other MCU with different level of power and hardware support. Whereas Teensy 4.1, operating at \texttt{600\ MHz}, is our primary development platform which we will evaluate closely, we aim to run our implementations on the Raspberry Pi Pico 2\footnote{\url{https://datasheets.raspberrypi.com/pico/pico-2-product-brief.pdf}}, pictured to the right in fig.~\ref{fig:mcu}, operating at \texttt{150\ MHz}. Due to the lower computational power of this MCU, there may be significant issues in utilizing this weaker model, yet it is significantly cheaper and easier to access. Testing will consist solely of running the implementation on these controllers and measuring execution speed and correctness of the output.

@@

The raw bits from the ADC can potentially have some deterministic patterns, and as such have to be processed somehow in order to remove these patterns. Several methods exist for this purpose, and for our work, we will perform this post-processing via Toeplitz extraction. The main focus of this study is to implement this extraction algorithm as effectively as possible on resource constrained hardware.

@@

@@

With the assumption that the OQRNG-device produces a truly random analog signal, we can now clearly define the scope in which this thesis operates. Considering the maximum conversion speed from the ADC and the USB-output from the MCU, we have a clear bound over 24 Mbit/s (\emph{imposed by the ADC}) in which Toeplitz extraction needs to be executed. Any speeds over 24 Mbit/s allows us to upgrade the ADC iteratively to continue increasing the output speed. Any implementation of Toeplitz extraction must then execute fast enough on any given microcontroller feasible for the proposed quantum RNG-thumbstick as to not be the decisive limiting factor.

@@

Another important point that is often neglected in randomness extraction using Toeplitz matrices is how the seed key is handled when forming the Toeplitz matrix. Numerous systems, such as those developed by Chouhan et al. \cite{toeplitz-desc} and Zhang et al. \cite{toeplitz}, depends on fixed seeds. Fixed seeds can create security risks over prolonged operation time. To tackle this problem, Lin et al \cite{lin} proposed a method for seed-renewable Toeplitz post-processing in QRNG. Their strategy incorporates a dynamic seed pool within the FPGA, where each instance of post-processing picks a new, randomly selected seed, minimizing temporal correlations between extractions. Furthermore, an external seed updating mechanism via Peripheral Component Interconnect Express (\emph{PCIe}) ensures that seeds are refreshed whenever a certain security limit is reached. Compared to fixed-seed methods, this renewable approach enhances cryptographic robustness and ensures sustained high-security randomness extraction in real-world applications.

Efficient Toeplitz matrix-vector multiplication (\emph{TMVM}) is critical for optimizing randomness extraction that relies on Toeplitz, particularly in constrained hardware environments. Liao et al. \cite{liao} showed that this process could be greatly accelerated using Fast Fourier Transform (\emph{FFT}) and its inverse (\emph{IFFT}) -- reducing computational complexity from \(O(n^2)\) to \(O(n \log n)\). Their implementation on FPGA utilized this approach for deep neural networks, resulting in a 28.7 times decrease in model size while still achieving fast inference speeds. By using FFT and IFFT acceleration, Toeplitz post-processing for randomness extraction could achieve higher throughput. This could potentially improve performance and reduce latency.

@@

\textbf{Iteration 3 - Batching and Hardware optimization:} This iteration focused on optimizing performance through batching and the use of ARM-native instructions, beginning by testing batching alone. This was followed by isolated use of ARM instructions such as \texttt{\_\_builtin\_popcountll()} (\emph{which counts the number of set bits in an unsigned integer}). After establishing their individual effect, we combined both techniques, Multiple batch sizes were tested to determine their impact. More details and benchmarks for each configuration can be found in Section 6.

@@

Phase two focused on addressing inefficiencies and design issues that were unintentionally introduced or not addressed during earlier iterations. Rather than continuing with new algorithmic ideas, this phase aimed to identify and fix structural problems identified during the previous phase. Several assumptions from phase one were re-evaluated -- such as the benefits of certain data structures or abstractions.

@@

The way \texttt{ent} calculates the entropy of a given binary file is by checking the frequency of certain byte values. If some byte values are more frequent, the binary file has a low entropy -- essentially signifying that there are predictable patterns. Conversely, if all byte values appear roughly equally, the binary file has a high entropy -- meaning that it is more random, and less predictable. The maximum entropy of any given file is \(\log_2(256) = 8 \text{ bits per byte}\), meaning that a perfectly random binary file has an entropy score of 8. For the purposes of our work, we consider values greater than \(7.99\) to be acceptable.

Furthermore, \texttt{ent} also provides a measurement of the possible compression ratio of the file, e.g.~how much more the file can be theoretically compressed. This is calculated as \(1 - \frac{\text{entropy}}{\text{maximum entropy}}\), signifying that the file cannot be compressed further -- that the data is already as compressed as its random distribution allows. Essentially, this informs us that the file has no more detectable statistical patterns or compressible regularities remaining. We consider any file where \texttt{ent} reports a compression ratio of \(0\%\) to be acceptable.

In regards to execution time, Hyncica et. al. \cite{micromeasurements} propose that measuring execution time of algorithms directly via the microcontrollers internal timers (\emph{while subtracting the interrupt overhead}) provides adequate measurements of the execution speed of an algorithm. An additional advantage is that the same code can be used to measure execution speed on several different microcontrollers, rather than relying on counting CPU cycles (\emph{as the process for this may vary greatly between controllers}).

@@

To isolate the execution speed of the algorithm as precisely as possible, we only intend to measure the execution speed of the algorithm. Whereas there is some other code present in each iteration (\emph{primarily packing the output data to be sent to the host computer in a workable format}), the brunt of the work should reasonably be performed by the algorithm. As such, we expect the time spent on executing other code to be negligible. Furthermore, this code executes far less often and does less work than the algorithm, and does not exponentially increase with larger the input and output requirements.

@@

@@

We created a script to facilitate easier testing, which is attached as an appendix to this paper. Using this script, all bitstrings used for testing can be evaluated against the baseline, ensuring that the output from the new iteration matches the baseline exactly. Furthermore, the script also provides the average execution time of only the Toeplitz extraction in microseconds. A brief overview of the architecture of this test script can be seen in Figure \ref{fig:testscript-architecture}.

@@

Table \ref{tab:iter0} presents the average execution time of the naive implementation (\emph{iteration 0}) across different input sizes for both the Teensy 4.1 and the Raspberry Pi Pico 2. The results reveal that the Teensy consistently achieves lower execution times compared to the Pico, with performance differences becoming more pronounced as the bit size increases. This behavior aligns with the expected difference in hardware specifications, such as CPU clock frequency and memory bandwidth.

@@

Additionally, the isolated effect of applying a single bitmask operation \texttt{\&\ 1} was evaluated in the naive implementation. This resulted in a slight reduction in average execution time, from \texttt{13.1564} \(\mu s\) to \texttt{13.1000} \(\mu s\), although this specific result is not included in the tables above.

@@

Table \ref{tab:iter3} presents the average execution time of iteration 3 on the Teensy 4.1 for varying batching and bit sizes. The notation \texttt{64/64}, \texttt{512/512} and \texttt{1024/1024} refers to the input bit size and batching size.

To evaluate the effect of batching granularity, additional tests were conducted using batch sizes of 32, 256 and 512 for various input sizes. The observed difference in execution time between low and high batching configurations ranged approximately from 10 to 50 microseconds, depending on the input size.

For comparative analysis, the same 1024/1024-bit configuration was executed on the Raspberry Pi Pico 2. The average execution time was measured at \texttt{80806.006} \(\mu s\) indicating a substantially lower throughput compared to the Teensy.

@@

Table \ref{tab:iter4} presents the execution time of iteration 4, which utilizes a 4x unrolled bit-processing loop. Compared to earlier iterations, this approach yields a substantial performance improvements on both the Teensy and the Raspberry Pi Pico MCUs.

@@

An additional fully unrolled variant, where loops were entirely eliminated, produced a measured execution time of \texttt{0.0491} \(\mu s\). However, this result was later deemed invalid due to packaging error in the implementation. Specifically, the algorithm was incorrectly designed to transmit only a single output bit instead of full 32-bit expected output. This led to an underestimation of the true execution time, as the cost of correctly formatting and sending the full result was omitted from measurement.

@@

Table \ref{tab:iter5} presents the results of iteration 5, in which the original vector structure was removed in favor of utilizing fixed-size integer types \texttt{uint32\_t} and \texttt{uint64\_t}. This iteration led to an extremely efficient implementation that approaches the physical execution limits of the Teensy 4.1. Due to the removal of \texttt{std::vector}, this implementation is limited by the available integer sizes of the MCU, and could as such only be tested with at most 64 bits. The ramifications of this will be discussed in more detail in Section 7. However, the results remain inconsistent -- while the average execution speed is low, individual measurements vary, indicating that the implementation does not yet achieve stable execution at the architectural limit.

@@

Table \ref{tab:iter6} presents the results of iteration 6 using 64 bit size, which was introduced to address the structural limitations encountered in iteration 5. Specifically, the fixed-width output constrains imposed by the removal of vector. Several alternative data structures were evaluated, with the original vector-based implementation from iteration 1 serving as a baseline.

Notably, while the bitset-based implementation achieved an average execution time of 0.0474\,\(\mu s\), this value represents an average and not a consistently stable result. Individual measurements showed significant variance. This suggests that, despite nearing the theoretical limits outlined in Section 3.3, the implementation has not yet reached a consistently minimal execution time. Additional tests at 128-bit input size yielded 179.8712\,\(\mu s\), confirming that this approach is not suitable for increasing the input and output bit sizes.

@@

Observing the results, we can place them into the context of the limits imposed by the ADC, as discussed in Section 3.2. In equation (5) (\emph{as seen in Section 5.3}), we calculate the average execution speed required to be \(2.667 \mu s\) for the ADC suggested in Section 3.2. From Tables \ref{tab:iter5} and \ref{tab:iter6} we can see execution speeds well below that, with iteration 5 as well as iteration 6 -- specifically with bitsets -- being the most prominent results to discuss. Here, our implementation achieved speeds of \(0.0501 \mu s\) and \(0.0474 \mu s\) respectively which within the bounds of the ADC listed in Section 3.2. The built in ADC on Teensy has an effective sample rate of 1 MSPS (\emph{e.g.~half of MAX11102AUB}), which would require an average execution speed of \textasciitilde{}\(2.667 \mu s / 2 \approx 1.334 \mu s\), making our implementation suitable even for that limited conversion speed.

@@

However, it is important to note that both of these implementations has a hard limit of working on 64 bits input. Whereas larger bitstrings still yield results with high entropy, the execution speed increases far above the hardware limitations, as evident in Table \ref{tab:iter6}.

One issue still remains concerning the size of the input string and subsequent output string. The larger the size of the matrix provided to the Toeplitz function (\emph{e.g.~the larger our input}), the higher the potential overall entropy becomes. Iteration 5 is hard capped at using only 64 bits input, as we need to store the bits in fixed-size integers before running them through the algorithm, and on embedded machines we have no larger integer types available, which may or may not prove to not be secure enough for use in real world applications. This, however, is out of scope for this particular thesis, but a consideration nonetheless.

@@

As discussed in Section 7, this comes at a price. The implementation that did indeed perform above expectation could only output a 32-bit integer, as we cannot take any larger input than 64 bits without any more complex data structure. How much of a flaw this is for a system utilizing this method of OQRNG is beyond the scope of this thesis to investigate. However, it is likely that a larger output is required for any application revolving around tight security. For less critical systems and less essential functionality, this method might, however, prove just secure enough. For instance, an IoT-device that requires a random number output for short-term symmetric cryptography when connecting to a user device may be an ideal candidate for this method of OQRNG.
